Converging High-Throughput and High-Performance Computing Paradigms: \\ A Case Study of Using Abstractions to Bridge the Divide

### 2017-03-17 overall.tex
Iteration and shortening IV and V; add new data from Sergey

Currently, ATLAS deploys 20 instances of PanDA Broker on 4 Titan's DTNs, 5
instances per DTN. Each broker submits and manages the execution of 15 to 300
detector simulation jobs, one job for each work node, and a theoretical maximum
concurrent use of 96,000 cores. Since November 2015, PanDA Brokers have operated
only in backfill mode, without a defined time allocation, and running at the
lowest priority on Titan.

We evaluate the efficiency, scalability and reliability of the PanDA Brokers
along two dimensions: (i) the amount of backfill availability utilized by the
PanDA Brokers; and (ii) the amount of runtime spent performing detector
simulations. We based these evaluations on measuring the number of cores
utilized by ATLAS on Titan, the overall backfill availability, the number of
detector simulations performed on all the resources available to ATLAS and on
Titan, and the failure rate of these simulations. All the measurements were
performed between January 2016 and February 2017.

@@

Figure~\ref{fig:core-hours-utilization} shows the number of core-hours used by
ATLAS on Titan, from January 2016 to February 2017. During that period, ATLAS
consumed a total of 73.8M core-hours, for an average of $\approx$ 7M core-hours
a month, with a minimum of 3.3M core-hours in April 2016 and a maximum 14.8M
core-hours in February 2017.

@@

Figure~\ref{fig:backfill-utilization} shows backfill utilization from January
2016 to February 2017 (hereafter called experiment time window).
\jhanote{Neither figure show efficiency. Both show utilization.} Efficiency is
defined as the fraction of Titan’s cores available via backfill that was
utilized by the PanDA Brokers. \jhanote{I would plot the efficiency on the Y2
axis for both.} The brokers reached 18\% average efficiency when measured across
the experiment time window, with a minimum 8.9\% efficiency on April 2016 and a
maximum 33.5\% efficiency on February 2017. The number of total backfill cores
available was 38.1M in April 2016, and 33.1M in February 2017. This shows that
the measured difference in efficiency did not depend on a comparable difference
in total backfill availability.

@@

Figure~\ref{fig:hpc-workload-utilization} shows that between January 2016 and
February 2017, about 2.25M detector simulation jobs were completed on Titan, for
a total of 225M events processed. This is equivalent to 0.9\% of all the 250M
detector simulations performed by ATLAS in the same period of time, and 3.5\% of
the 6.6B events processed by those jobs. Comparatively, Titan contributed 3.9\%
of the total of around 200K cores available to ATLAS on Grid, Cloud, and HPC
infrastructures together. These figures confirms the relevance of
supercomputers' resource contribution to the LHC Run 3, especially when
accounting for the amount of unused backfill availability and the rate of
improvement of PanDA efficiency.

@@

On February 2017, PanDA Brokers used almost twice as much backfill availability
than in any other month. No relevant code update was made during that period and
logs indicated that the brokers were able to respond more promptly to backfill
availability. This is likely due to hardware upgrades on the DTNs. The absence
of continuous monitoring of the nodes does not allow to quantify bottlenecks but
spot measurements of the DTNs load indicate that a faster CPU and better
networking were likely responsible for the improved performance.

@@

All the detector simulations executed on Titan process 100 events. This number
of events is consistent with the physics of the use case and with the average
duration of backfill availability. The duration of a detector simulation is a
function of the number of events: the more events, the longer the simulation.
Not all events take the same time to be simulated:
Figure~\ref{fig:100event-distrib} shows the distribution of that duration for
XXXXX events simulated on Titan. When compared to the distribution of backfill
availability on Titan in Figure~\ref{fig:backfill-distrib}, 100 events appear to
be a good compromise.

\begin{figure}[htp]
    \includegraphics[clip,width=\columnwidth]{figures/tx_simulation_distribution.pdf}
\caption{Distribution of the time taken by a Geant4 detector simulation to simulate one event. 3000 Events; 30 Titan work nodes; 16 simulations per node; 100 events per node. Average time per event \~14 min. Broad distribution from \~1 to \~40 minutes.}
\label{fig:100event-distrib}
\end{figure}

\begin{figure}[htp]
    \includegraphics[clip,width=\columnwidth]{figures/tx_simulation_distribution.pdf}
\caption{\mtnote{Placeholder for the distribution of backfill walltime availability on Titan.}}
\label{fig:backfill-distrib}
\end{figure}

Usually, detector simulations performed on the Grid process a higher number of
events than those on Titan. This explains the difference between the percentage
of detector simulations (9\%) and of events computed by those simulations
(3.9\%) on Titan. PanDA Broker could fit the number of events to the walltime of
the backfill availability on the base of the distribution of
Figure~\ref{fig:backfill-distrib}. That specific number of event could then be
pulled from the PanDA Event service~\cite{panda-event-service} and given as
input to one or more simulations. Once packaged into a MPI PBS job, these
simulations would better fit backfill availability, increasing the efficiency of
PanDA Brokers.

The transition from a homogeneous to a heterogeneous number of events per
detector simulation has implications for the application layer. An even number
of events across simulations makes it easier to partition, track and package
events across simulations, especially when they are performed on both the Grid
and Titan. A homogeneous number of events also helps to keep the size and
duration of other stages of the ATLAS MC workflow (\S\ref{ssec:panda-titan})
more uniform. Further analysis is needed to evaluate the trade offs between
increased efficiency of resource utilization and the complexity that would be
introduced at the application layer.

@@

% The number of concurrent detector simulation jobs that PanDA Broker can manage
% depends on the DTNs' resources. Each PanDA Broker can execute 1 MPI PBS job on
% up to 300 work nodes for a total of 4,800 cores. Scaling above this threshold
% requires instantiating multiple PanDA Brokers on the same DTN and, once the
% DTN resources are saturated, on multiple DTNs.

@@

\begin{itemize}
    \item problem with lustre addressed with ramdisk
    \item Competition for cache in AMD processors.
    \item Summary overall figure of performance analogous to those produced with NGE. Diagram as discussed in F2F meeting at Rutgers.
    \item Comparison hardware performance Titan/Grid
\end{itemize}

@@

Initially, all the shared libraries of AthenaMP and the python scripts for the
configuration stage were stored on the Spider 2 Lustre file system. However, the
I/O patterns of the initialization and configuration stages degraded the
performance of the filesystem (Figure~\ref{}\mtnote{We may want to aggregate
some of the diagrams about lustre's experiments here}). Since Spider 2 is a
center-wide file system, this resulted in  performance degradation for all OLCF
resources and users. \mtnote{I am afraid the details about the trace are too
specific given the space constraints of a SC submission. Please feel free to
uncomment it if you disagree.}
% As can be seen in Listing~\ref{mdstrace} the metadata I/O activity for ATLAS
% exhibits a spike corresponding to the beginning of the runs before tapering
% off.

@@

After careful characterization of the impact of ATLAS jobs on Lustre, the
performance degradation was addressed by moving the AthenaMP distribution to a
read-only NFS directory, shared among DTNs and work nodes. This eliminated the
problem of metadata contention, improving metadata read performance from \~6,300
seconds on Lustre to \~1,500 seconds on NFS. Further, at configuration time, the
input files of each detector simulation job were stored to a ramdisk on the work
nodes. This offered a marked improvement of reading time: from 1,320 seconds on
Lustre to 40 seconds on NFS.

@@

The AMD Opteron 6274 CPU used on Titan has 16 cores, divided into 8 compute
units. Each compute units has 1 floating point (FP) scheduler shared between 2
cores. When using 16 cores for FP-intensive calculations, each pair of cores
competes for a single FP scheduler. This creates the overhead shown in
Figure~\ref{fig:comparison-8-16cores}: the mean runtime per event for 8
concurrent simulations computing 50 events is 10.8 minutes, while for 16
simulations is 14.25 minutes (consistent with
Figure~\ref{fig:100event-distrib}). The allocation policy based on work nodes
and not cores used, does not justify the use of 1/2 of the cores available.

\begin{figure}[htp]
    \includegraphics[clip,width=\columnwidth]{figures/tx8_tx16_comparison.pdf}
\caption{Comparison between distributions of the time taken by a Geant4 detector simulation to simulate one event when placing 2 simulations (h1) or 1 simulation (h2) per CPU. 2 simulation use 16 cores per node, 1 simulation 8, 1 per compute unit. 50 Events; 1 Titan work nodes; 16 work threads per node; 100 events per node. Average time per event \~14 min. Broad distribution from \~1 to \~40 minutes.}
\label{fig:comparison-8-16cores}
\end{figure}

The compute performance of Titan for detector simulation jobs is, on average,
lower that that of LCWG. Figure~\ref{}\mtnote{Sergey: I remember you showed us
some plots comparing Geant4 simulation's performance on Titan and some of the
Grid sites. Could we use one or two of the here?} shows the difference in
execution time of the same type of simulation between Titan and
\ldots\mtnote{Add the name of the grid sites here}. This is not an issue with
the design of PanDA Brokers and, more in general, with the migration of part of
the ATLAS workload to supercomputers: Titan has been active since 2012 and more
modern CPUs are expected to offer better performance. A pilot project is already
in place to test PanDA Broker on Summit, the next generation supercomputer that
will become active in 2018 at OLCF.

>>>>

Currently, ATLAS deploys 20 instances of PanDA Broker on 4 Titan's DTNs, 5
instances per DTN. Each broker submits and manages the execution of 15 to 300
ATLAS jobs, one job for each worker node, and a theoretical maximum concurrent
use of 96,000 cores. Since November 2015, PanDA Brokers have operated only in
backfill mode, without a defined time allocation, and running at the lowest
priority on Titan.

We evaluate the efficiency, scalability and reliability of the PanDA Brokers
along two dimensions: (i) the amount of backfill availability utilized by the
PanDA Brokers; and (ii) the amount of runtime spent performing detector
simulations. We based these evaluations on measuring the number of cores
utilized by ATLAS on Titan, the overall backfill availability, the number of
detector simulations performed on all the resources available to ATLAS and on
Titan, and the failure rate of these simulations. All the measurements were
performed between January 2016 and February 2017, hereafter called `experiment
time window'.

@@

Figure~\ref{fig:core-hours-utilization} shows the number of core-hours used by
ATLAS on Titan during the experiment time window. ATLAS consumed a total of
73.8M core-hours, for an average of $\approx$ 7M core-hours a month, with a
minimum of 3.3M core-hours in April 2016 and a maximum 14.8M core-hours in
February 2017.

@@

Figure~\ref{fig:backfill-utilization} shows backfill utilization during the
experiment time window.\jhanote{Neither figure show efficiency. Both show
utilization.} Efficiency is defined as the fraction of Titan’s cores available
via backfill that was utilized by the PanDA Brokers. \jhanote{I would plot the
efficiency on the Y2 axis for both.} The brokers reached 18\% average
efficiency, with a minimum 8.9\% efficiency on April 2016 and a maximum 33.5\%
efficiency on February 2017. The number of total backfill cores available was
38.1M in April 2016, and 33.1M in February 2017. This shows that the measured
difference in efficiency did not depend on a comparable difference in total
backfill availability.

@@

Figure~\ref{fig:hpc-workload-utilization} shows that during the experiment time
window, about 2.25M detector simulation jobs were completed on Titan, for a
total of 225M events processed. This is equivalent to 0.9\% of all the 250M
detector simulations performed by ATLAS in the same period of time, and 3.5\% of
the 6.6B events processed by those jobs. Comparatively, Titan contributed 3.9\%
of the total of around 200K cores available to ATLAS on Grid, Cloud, and HPC
infrastructures together. These figures confirms the relevance of
supercomputers' resource contribution to the LHC Run 3, especially when
accounting for the amount of unused backfill availability and the rate of
improvement of PanDA efficiency.

@@

On February 2017, PanDA Brokers used almost twice as much backfill availability
than in any other month. No relevant code update was made during that period and
logs indicated that the brokers were able to respond more promptly to backfill
availability. This is likely due to hardware upgrades on the DTNs. The absence
of continuous monitoring of those nodes does not allow to quantify bottlenecks
but spot measurements of their load indicate that a faster CPU and better
networking were likely responsible for the improved performance.

@@

Every detector simulation executed on Titan process 100 events. This number of
events is consistent with the physics of the use case and with the average
duration of backfill availability. The duration of a detector simulation is a
function of the number of events: the more events, the longer the simulation.
Not all events take the same time to be simulated:
Figure~\ref{fig:100event-distrib} shows a distribution of the simulation time of
1 event from $\approx$ 2 to $\approx$ 40 minutes, with a mean of $\approx$ 14
minutes. Considering that each worker node process up to 16 events concurrently,
100 events takes around 2 hours to process.

\begin{figure}[htp]
    \includegraphics[clip,width=\columnwidth]{figures/tx_simulation_distribution.pdf}
\caption{Distribution of the time taken by a Geant4 detector simulation to simulate one event. 3000 Events; 30 Titan worker nodes; 16 simulations per node; 100 events per node. Average time per event $\approx$ 14 min. Broad distribution from $\approx$ 2 to $\approx$ 40 minutes.}
\label{fig:100event-distrib}
\end{figure}

Figure~\ref{fig:backfill-distrib} shows the distribution of backfill
availability on Titan as a function of number of nodes and the time of their
availability (i.e., walltime). The mean number of nodes over the experiment
window time was $\approx$ 267, and their mean walltime was 1445 minutes.
Detector simulations of 100 events, enable to use down to 1 node for 1/4 of the
mean walltime of backfill availability. As such, it offers a good compromise for
the PanDA Broker efficiency.

\begin{figure}[htp]
    \includegraphics[clip,width=\columnwidth]{figures/avail_time_vs_nodes_Oct_28_Nov_3_2015.png}
\caption{\mtnote{Waiting to know whether we have data for the experiment time window.}}
\label{fig:backfill-distrib}
\end{figure}

Usually, detector simulations performed on the Grid process 10 times the number
of events than those on Titan. This explains the difference between the
percentage of detector simulations (9\%) and of events computed by those
simulations (3.9\%) on Titan. PanDA Broker could fit the number of events to the
walltime of the backfill availability on the base of the distribution of
Figure~\ref{fig:backfill-distrib}. That specific number of event could then be
pulled from the PanDA Event service~\cite{panda-event-service} and given as
input to one or more simulations. Once packaged into the MPI script submitted to
titan's PBS batch system, these simulations would better fit backfill
availability, increasing the efficiency of PanDA Brokers.

The transition from a homogeneous to a heterogeneous number of events per
detector simulation has implications for the application layer. An even number
of events across simulations makes it easier to partition, track and package
events across simulations, especially when they are performed on both the Grid
and Titan. A homogeneous number of events also helps to keep the size and
duration of other stages of the MC workflow (\S\ref{ssec:panda-titan}) more
uniform. Further analysis is needed to evaluate the trade offs between increased
efficiency of resource utilization and the complexity that would be introduced
at the application layer.

@@

% The number of concurrent detector simulation jobs that PanDA Broker can manage
% depends on the DTNs' resources. Each PanDA Broker can execute 1 MPI PBS job on
% up to 300 worker nodes for a total of 4,800 cores. Scaling above this
% threshold requires instantiating multiple PanDA Brokers on the same DTN and,
% once the DTN resources are saturated, on multiple DTNs.

@@

\begin{itemize}
    \item problem with lustre addressed with ramdisk (done)
    \item Competition for cache in AMD processors. (done)
    \item Comparison hardware performance Titan/Grid (data acquired)
    \item Summary overall figure of performance analogous to those produced with NGE. Diagram as discussed in F2F meeting at Rutgers. (waiting)
\end{itemize}

@@

Initially, all the shared libraries of AthenaMP and the python scripts for the
configuration stage were stored on the Spider 2 Lustre file system. However, the
I/O patterns of the initialization and configuration stages degraded the
performance of the filesystem (Figure~\ref{}\mtnote{We may want to aggregate
some of the diagrams about lustre's experiments here}). Since Spider 2 is a
center-wide file system, this resulted in  performance degradation for all OLCF
resources and users. \mtnote{I am afraid the details about the trace are too
specific given the space constraints of a SC submission. Please feel free to
uncomment it if you disagree.}

% As can be seen in Listing~\ref{mdstrace} the metadata I/O activity for ATLAS
% exhibits a spike corresponding to the beginning of the runs before tapering
% off.

@@

After careful characterization of the impact of ATLAS jobs on Lustre, the
performance degradation was addressed by moving the AthenaMP distribution to a
read-only NFS directory, shared among DTNs and worker nodes. This eliminated the
problem of metadata contention, improving metadata read performance from
$\approx$ 6,300 seconds on Lustre to $\approx$ 1,500 seconds on NFS. Further, at
configuration time, the input files of each detector simulation job were stored
to a ramdisk on the work nodes. This offered a marked improvement of reading
time: from 1,320 seconds on Lustre to 40 seconds on NFS.

@@

The AMD Opteron 6274 CPU used on Titan has 16 cores, divided into 8 compute
units. Each compute units has 1 floating point (FP) scheduler shared between 2
cores. When using 16 cores for FP-intensive calculations, each pair of cores
competes for a single FP scheduler. This creates the overhead shown in
Figure~\ref{fig:comparison-8-16cores}: the mean runtime per event for 8
concurrent simulations computing 50 events is 10.8 minutes, while for 16
simulations is 14.25 minutes (consistent with
Figure~\ref{fig:100event-distrib}). Despite an inefficiency of almost 30\%,
Titan's allocation policy based on number of worker nodes used instead of number
of cores does not justify the use of 1/2 of the cores available.

\begin{figure}[htp]
    \includegraphics[clip,width=\columnwidth]{figures/tx8_tx16_comparison.pdf}
    \caption{Comparison between distributions of the time taken by a Geant4
    detector simulation to simulate one event when placing 2 simulations (h1) or
    1 simulation (h2) per CPU. 2 simulation use 16 cores per node, 1 simulation
    8, 1 per compute unit. 50 Events; 1 Titan worker nodes; 16 work threads per
    node; 100 events per node.}
\label{fig:comparison-8-16cores}
\end{figure}

The compute performance of Titan for detector simulation jobs is, on average,
lower that that of LCWG. Figure~\ref{fig:comparison-titan-grid} shows the
difference in execution time of the same type of simulation between Titan and
\ldots\mtnote{Add the name of the grid sites here}. This is not an issue with
the design of PanDA Brokers and, more in general, with the migration of part of
the ATLAS workload to supercomputers: Titan has been active since 2012 and more
modern CPUs are expected to offer better performance. A pilot project is already
in place to test PanDA Broker on Summit, the next generation supercomputer that
will become active in 2018 at OLCF.

\begin{figure}[htp]
    \includegraphics[clip,width=\columnwidth]{figures/tx8_tx16_comparison.pdf}
    \caption{}
\label{fig:comparison-titan-grid}
\end{figure}
